Hybrid Threads for the Parrot Virtual Machine

### 2012-03-08 Paper.tex
s/explicitely/explicitly/ and s/implicitely/implicitly/

The Perl 6 design process began with asking the Perl users for what they were expecting from the new version of the language. The very first feature that got asked for was well integrated multithreading support\cite{RFC1}. Perl 5 had two different implementations of thread support. In the first model, called 5005threads, data was shared by default and shared access to data had to be explicitely synchronized. This was similiar to the models used by languages such as C or Java. The implementation however suffered from data corruption and crashes and thus was not recommended for production use\cite{ThreadManual}. Perl 5.6 introduced the newer model called ithreads, mostly as a way to emulate fork() on Win32 platforms. Perl 5.8 exposed this API (Application Programming Interface) to the user of the programming language. In this new model, all data is copied to each thread and afterwards thread local. Data must be explicitely shared between threads. In other words, in Perl threads are not lightweight at all. They have severe impact on memory usage, writes to shared data are expensive and still not all features of the language are usable in threaded programs.

@@

In Java responsibility for preventing concurrency issues is obliged on the user code. The language provides synchronization primitives like mutexes, but the interpreter (the Java Virtual Machine, JVM) does not protect itself. The class library provides the user with high level data structures explicitely designed for multithreaded scenarios.

@@

The CPython implementation of the Python runtime uses a Global Interpreter Lock (GIL) to protect its internal consistency\cite{PythonThreadingManual}. This is a single lock taken whenever the interpreter executes Python bytecode. Because of this lock, only one thread can execute bytecode at any time so all built-in types and the object model are implicitely type safe. The drawback is that Python code cannot benefit from having multiple CPU cores available. However I/O operations and calls to external libraries are done without holding the GIL, so in applications with multiple I/O bound threads, there may still be a performance benefit from using multithreading.

@@

Contrary to other widely used virtual machines like the JVM or the CLR which are stack based, Parrot mirrors contemporary hardware CPUs more closely by being register based. A stack based virtual machine usually fetches the operands of an operation from the top of a stack and puts the result back on top. Thus the operands chosen implicitely by ordering of the operations allowing the opcodes to be quite small. In a register based VM on the other hand each operation has to specify the operands explicitely. Compilers for stack machines are simpler because they do not have to care about register allocations and code is independent of prior or subsequent code\cite{VMShowdown}. The rationale behind giving up the simplicity of a stack based implementation is the hope of simplifying just in time compilation and improved performance of nested function and method calls.

>>>>

The Perl 6 design process began with asking the Perl users for what they were expecting from the new version of the language. The very first feature that got asked for was well integrated multithreading support\cite{RFC1}. Perl 5 had two different implementations of thread support. In the first model, called 5005threads, data was shared by default and shared access to data had to be explicitly synchronized. This was similiar to the models used by languages such as C or Java. The implementation however suffered from data corruption and crashes and thus was not recommended for production use\cite{ThreadManual}. Perl 5.6 introduced the newer model called ithreads, mostly as a way to emulate fork() on Win32 platforms. Perl 5.8 exposed this API (Application Programming Interface) to the user of the programming language. In this new model, all data is copied to each thread and afterwards thread local. Data must be explicitly shared between threads. In other words, in Perl threads are not lightweight at all. They have severe impact on memory usage, writes to shared data are expensive and still not all features of the language are usable in threaded programs.

@@

In Java responsibility for preventing concurrency issues is obliged on the user code. The language provides synchronization primitives like mutexes, but the interpreter (the Java Virtual Machine, JVM) does not protect itself. The class library provides the user with high level data structures explicitly designed for multithreaded scenarios.

@@

The CPython implementation of the Python runtime uses a Global Interpreter Lock (GIL) to protect its internal consistency\cite{PythonThreadingManual}. This is a single lock taken whenever the interpreter executes Python bytecode. Because of this lock, only one thread can execute bytecode at any time so all built-in types and the object model are implicitly type safe. The drawback is that Python code cannot benefit from having multiple CPU cores available. However I/O operations and calls to external libraries are done without holding the GIL, so in applications with multiple I/O bound threads, there may still be a performance benefit from using multithreading.

@@

Contrary to other widely used virtual machines like the JVM or the CLR which are stack based, Parrot mirrors contemporary hardware CPUs more closely by being register based. A stack based virtual machine usually fetches the operands of an operation from the top of a stack and puts the result back on top. Thus the operands chosen implicitly by ordering of the operations allowing the opcodes to be quite small. In a register based VM on the other hand each operation has to specify the operands explicitly. Compilers for stack machines are simpler because they do not have to care about register allocations and code is independent of prior or subsequent code\cite{VMShowdown}. The rationale behind giving up the simplicity of a stack based implementation is the hope of simplifying just in time compilation and improved performance of nested function and method calls.
